Medical Training for Abortion Procedures and Physician Provision of Abortions in the United States

### 2016-04-30 paper.tex
Final round of edits from third reviewer

% Background on abortion in the United States
% Contextualizing these providers of abortion
The landscape of abortion in the United States is fairly evident in its imposition of restrictions of access to procedures for patients, but barriers also exist for providers.
Abortion is a common operation in the United States, with an estimated 1.1 million abortions performed in 2011. \autocite[pg. 7]{jones_abortion_2014}
However, geographic access to abortion is not uniform: in 2011, 89\% of US counties had no abortion or nonspecialized clinic with 38\% of women aged 15--44 lived in those counties, essentially unchanged from the geographic availability in 2008. \autocite[pg. 7]{jones_abortion_2014}
More specifically, the division of abortion providers is distinct between urban centers and rural areas.
While abortion providers are more concentrated in cities, in 2008, 69\% of counties in metropolitan areas lacked a provider while 97\% of nonmetropolitan counties lacked a provider. \autocite[pg. 46]{jones_abortion_2011}

The geographic disparity in access, however, does not correspond to a disparity in demand for abortion services.
A 2008--9 survey found that 97\% of obstetrician-gynecologists (OB/GYNs) in the United States under the age of 65 encountered patients seeking abortions while only 14.4\% of those practicing OB/GYNs provided abortions themselves. \autocite[pg. 611]{stulberg_abortion_2011}
In the United States, some 20,090 OB/GYNs were employed in May 2015, primarily in physicians' offices and outpatient care centers. \autocite{occupational_employment_and_wages_may_2015}
The primary places of employment are not primary centers for the provision of abortion services, though.
While 19\% of abortion providers in 2008 were physicians' offices, they only provided 1\% of all abortions at the greatest expense and at a low rate per location. \autocite[pg. 46--7]{jones_abortion_2011}
In comparison to the number of abortions in the United States, 32.8\% of births were Caesarean deliveries, about 1.3 million births in 2011. \autocite[pg. 5]{martin_births_2013}
Even though a Caesarean is a more complex procedure than an abortion, it is distinctly more accepted and has not generated the same controversy, despite its higher frequency.

@@

While abortion has had a long history \autocite[For a thesis about training for both procedural and elective abortions in the United States from 1920 to 2007, see:][]{ayres_providing_2009} in the United States, I will discuss more contemporary changes in training and practice.
% Changing environment of abortion healthcare
In 1973, the decision of \textit{Roe~v~Wade} enabled abortions to shift out of hospitals and into non-hospital clinics, thus generally removing the common procedure and its associated curricula from the US medical educational environment. \autocite[pg. 404]{aksel_unintended_2013}
Although the shift to abortion clinics expanded access to elective abortions because clinics added to the possible provider locations, the overall impact on female health had negative impacts in the long term.
Even though the media wondered if abortion should have been a solved issue just 10 years after \textit{Roe}\autocite{mark_obenhaus_abortion_1983}, it is evident that acceptance of abortion as an acceptable, safe, and routine medical operation still has not happened.
Another ten years later, in the early 1990s, there was a distinct lack of training medical school training in family planning, including abortion, but in 1994 Jody Steinauer started the meetings with fellow medical students that became the organization Medical Students for Choice. \autocite[pg. 404--5]{aksel_unintended_2013}

@@

% Enablers, barriers, consequences
Creating, maintaining, and teaching such a family planning and abortion program in medical residency is not as simple as following the mandate. Within OB/GYN residency programs that include abortion training, there are a number of factors that enable the training and others that are barriers to effective training.
Enablers of training, identified in a qualitative survey of participants in OB/GYN residency in New York City (NYC), include: program/facility location, access to high-capacity services, faculty commitment to training, Ryan Program \autocite{bixby_center_for_global_reproductive_health_education_????} \autocite{bixby_center_for_global_reproductive_health_home_????} \autocite{steinauer_benefits_2013} support, and interest of residents. \autocite[pg. 240--2]{guiahi_enablers_2013}
Identified barriers to training include: poor continuity in leadership, conflict between leaders, lacking exposure to second-trimester abortion, difficulties acquiring mifepristone, burden and incompleteness of opt-in training, and antiabortion values of leadership/staff. \autocite[pg. 242]{guiahi_enablers_2013}
The author emphasizes that these results are not generalizable due to the high-density, urban nature of the survey and the ``politically favorable environment'' experienced in NYC. \autocite[pg. 243]{guiahi_enablers_2013}
In other areas that are less accepting of providing abortion care, it is likely that the enabling factors will be less present and the barrier factors may be more significant or other barriers may be present.

The significance of the burden caused by opt-in abortion training noted in the survey by \citeauthor{guiahi_enablers_2013} is identified elsewhere by the practical arrangements required.
Nationally, elective abortion training in the United States was most frequency provided (for 48\% of residents surveyed in 2010) in freestanding clinics outside of the teaching hospital attended by the residents. \autocite[pg. 274]{turk_availability_2014} This indicates the continued separation of both the procedure and training of abortion from the usual domain of hospital-based medicine. Incomplete integration into education is an unfortunate loss for the OB/GYN residents.
While training in therapeutic abortion was available for the vast majority of residents offered either opt-in or no training in elective abortion, surveyed residents offered opt-in elective abortion training had similar experiences in counseling and indicated similar counseling and procedural skills as residents at programs with no elective abortion training. \autocite[pg. 275--6]{turk_availability_2014}
The relative lack of a positive result from opt-in training is not a reason to curtail abortion training, but an indication that comprehensive training is required to meet the intended results of the ACGME mandate on access to abortion training.

@@

Beyond peers, larger and more formalized organizational policies prevent some physicians from providing abortions.
The United States is currently in a transitional period of physicians' attitudes towards abortions.
From a survey in 2008--9, younger (26--35 years old) and near-retirement (56--65 years old) OB/GYNs were more likely to provide abortions than middle age (36--45; 46--55 years old) cohorts, with the younger cohort being most likely to provide abortions. \autocite[pg. 611]{stulberg_abortion_2011}
However, these younger physicians will experience barriers to providing abortion not just from their peers, but also from the organizations that employ them.
Catholic health networks can prevent private, independent practices using the network's buildings from providing abortions while nonsectarian health maintenance organizations may mandate abortion referrals for economic reasons rather than providing in-network care. \autocite[pg. 149]{freedman_obstacles_2010}
This religious-based policy might be less harmful to access if Catholic healthcare organizations were less prevalent, but the presence of Catholic healthcare in the United States is significant.
The Catholic Health Association of the United States states that Catholic-affiliated providers are the ``nation’s largest group of not-for-profit health care providers,'' are present in all 50 states, and that ``one-in-six patients in the U.S. is cared for in a Catholic hospital.''  \autocite{the_catholic_health_association_of_the_united_states_catholic_2016}
Working primarily in a Catholic medical facility is strongly associated with decreased likelihood of performing abortions, while facilities of other religious dominations did not have a significantly different likelihood of performing abortions compared to non-religious facilities. \autocite[pg. 612]{stulberg_abortion_2011}
Both the relationships between physicians and their peers and physicians and the policies of related healthcare organizations prevent physicians from making the individual choices in care they are qualified to choose to offer their patiens.

>>>>

% Background on abortion in the United States
% Contextualizing these providers of abortion
The landscape of abortion in the United States is fairly evident in its imposition of restrictions of access to procedures for patients, but barriers also exist for providers.
Abortion is a common operation in the United States, with an estimated 1.1 million abortions performed in 2011. \autocite[pg. 7]{jones_abortion_2014}
However, geographic access to abortion is not uniform: in 2011, 89\% of US counties had no abortion or nonspecialized clinic with 38\% of women aged 15--44 living in those counties, essentially unchanged from the geographic availability in 2008. \autocite[pg. 7]{jones_abortion_2014}
More specifically, the division of abortion providers is distinct between urban centers and rural areas.
While abortion providers are more concentrated in cities, in 2008, only 69\% of counties in metropolitan areas lacked a provider while 97\% of nonmetropolitan counties lacked a provider. \autocite[pg. 46]{jones_abortion_2011}

The geographic disparity in access, however, does not correspond to a disparity in demand for abortion services.
A 2008--9 survey found that 97\% of obstetrician-gynecologists (OB/GYNs) in the United States under the age of 65 encountered patients seeking abortions while only 14.4\% of those practicing OB/GYNs provided abortions themselves. \autocite[pg. 611]{stulberg_abortion_2011}
In the United States, some 20,090 OB/GYNs were employed in May 2015, primarily in physicians' offices and outpatient care centers. \autocite{occupational_employment_and_wages_may_2015}
The primary places of employment are not primary centers for the provision of abortion services, though.
While 19\% of abortion providers in 2008 were physicians' offices, they only provided 1\% of all abortions at the greatest expense and at a low rate per location. \autocite[pg. 46--7]{jones_abortion_2011}
In comparison to the number of abortions in the United States, 32.8\% of births were Caesarean deliveries, about 1.3 million of the births in 2011. \autocite[pg. 5]{martin_births_2013}
Even though a Caesarean is a more complex procedure than an abortion, it is distinctly more accepted than and has not generated the same controversy as abortion, despite its higher frequency.

@@

While abortion has had a long history \autocite[For a thesis about training for both procedural and elective abortions in the United States from 1920 to 2007, see:][]{ayres_providing_2009} in the United States, I will discuss contemporary changes in training and practice.
% Changing environment of abortion healthcare
In 1973, one of the results of \textit{Roe~v~Wade} was allowing abortion procedures to shift out of hospitals and into non-hospital clinics, thus generally removing the common procedure and its associated curricula from the US medical educational environment. \autocite[pg. 404]{aksel_unintended_2013}
Although the shift to abortion clinics expanded access to elective abortions because clinics added to the possible provider locations, the overall impact on female health had negative impacts in the long term.
Even though the media opined that abortion should have been a solved issue just 10 years after \textit{Roe}\autocite{mark_obenhaus_abortion_1983}, it is evident that acceptance of abortion as an acceptable, safe, and routine medical operation still has not happened.
Another ten years later, in the early 1990s, there was a distinct lack of training medical school training in family planning, including abortion, but in 1994 Jody Steinauer started the meetings with fellow medical students that became the organization Medical Students for Choice. \autocite[pg. 404--5]{aksel_unintended_2013}

@@

% Enablers, barriers, consequences
Creating, maintaining, and teaching such a family planning and abortion program in medical residency is not as simple as following the mandate. Within OB/GYN residency programs that include abortion training, there are a number of factors that enable the training and others that are barriers to effective training.
Enablers of training, identified in a qualitative survey of participants in OB/GYN residency in New York City (NYC), include: program/facility location, access to high-capacity services, faculty commitment to training, Ryan Program \autocite{bixby_center_for_global_reproductive_health_education_????} \autocite{bixby_center_for_global_reproductive_health_home_????} \autocite{steinauer_benefits_2013} support, and interest of residents. \autocite[pg. 240--2]{guiahi_enablers_2013}
Identified barriers to training include: poor continuity in leadership, conflict between leaders, lack of exposure to second-trimester abortion, difficulty acquiring mifepristone, \note{One of the drugs used to induce medical abortion, usually with misoprostol.} burden and incompleteness of opt-in training, and antiabortion values of leadership/staff. \autocite[pg. 242]{guiahi_enablers_2013}
The author emphasizes that these results are not generalizable due to the high-density, urban nature of the survey and the ``politically favorable environment'' experienced in NYC. \autocite[pg. 243]{guiahi_enablers_2013}
In other areas that are less accepting of providing abortion care, it is likely that the enabling factors will be less present and the barrier factors may be more significant or other barriers may be present.

The significance of the burden caused by opt-in abortion training noted in the survey by \citeauthor{guiahi_enablers_2013} is identified elsewhere by the practical arrangements required.
Nationally, elective abortion training in the United States was most frequency provided (for 48\% of residents surveyed in 2010) in freestanding clinics outside of the teaching hospital attended by the residents. \autocite[pg. 274]{turk_availability_2014} This indicates the continued separation of both the procedure and training of abortion from the usual domain of hospital-based medicine. Incomplete integration into education is an unfortunate loss for the OB/GYN residents.
While training in therapeutic abortion was available for the vast majority of residents offered either opt-in or no training in elective abortion, surveyed residents offered opt-in elective abortion training had similar experiences in counseling and indicated similar counseling and procedural skills as residents at programs with no elective abortion training. \autocite[pg. 275--6]{turk_availability_2014}
The matching educational results from opt-in elective abortion training as compared to no available training is not a reason to curtail abortion training, but an indication that comprehensive training is required to meet the intended results of the ACGME mandate on access to abortion training.

@@

Beyond peers, larger and more formalized organizational policies prevent some physicians from providing abortions.
The United States is currently in a transitional period of physicians' attitudes towards abortions.
From a survey in 2008--9, younger (26--35 years old) and near-retirement (56--65 years old) OB/GYNs were more likely to provide abortions than middle age (36--45; 46--55 years old) cohorts, with the younger cohort being most likely to provide abortions. \autocite[pg. 611]{stulberg_abortion_2011}
However, these younger physicians will experience barriers to providing abortion not just from their peers, but also from the organizations that employ them.
Catholic health networks can prevent private, independent practices using the network's buildings from providing abortions while nonsectarian health maintenance organizations may mandate abortion referrals for economic reasons rather than providing in-network care. \autocite[pg. 149]{freedman_obstacles_2010}
This religious-based policy might be less harmful to access if Catholic healthcare organizations were less prevalent, but the presence of Catholic healthcare in the United States is significant.
The Catholic Health Association of the United States states that Catholic-affiliated providers are the ``nation’s largest group of not-for-profit health care providers,'' are present in all 50 states, and that ``one-in-six patients in the U.S. is cared for in a Catholic hospital.''  \autocite{the_catholic_health_association_of_the_united_states_catholic_2016}
Working primarily in a Catholic medical facility is strongly associated with decreased likelihood of performing abortions, while facilities of other religious dominations did not have a significantly different likelihood of performing abortions compared to non-religious facilities. \autocite[pg. 612]{stulberg_abortion_2011}
Both the relationships between physicians and their peers and physicians and the policies of related healthcare organizations prevent physicians from making the individual choices in care they are qualified to choose to offer their patients.
